---
title: "Introduction to ctreeMI: Conditional Inference Trees with Multiple Imputation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ctreeMI: Conditional Inference Trees with Multiple Imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
```

## Overview

`ctreeMI` implements the **stacked-imputation / Stack ÷ M** workflow for
conditional inference trees, as described in Sherlock et al. (2026). It
provides a simple, principled way to apply `ctree` when your data contain
missing values.

## Background

### Why ctree?

Conditional inference trees [@Hothorn2006] partition a dataset recursively
into subgroups (terminal nodes) using a permutation-based statistical
significance criterion. Unlike CART-style trees, ctree selects splits using
hypothesis tests with Bonferroni correction, which controls the probability
of spurious splits. This makes ctree well-suited for **exploratory,
person-centered analyses** in complex, high-dimensional data where
interactions are expected but cannot be specified *a priori*.

### The missing-data problem

Multiple imputation (MI) is the recommended approach for handling missing
data [@Rubin1987]. For models that produce scalar summaries (regression
coefficients, odds ratios), Rubin's rules provide a straightforward pooling
strategy. However, **pooling conditional inference trees is not
straightforward**: even small differences in imputed values across datasets
can produce structurally different trees — different root splits, different
variable orderings, different subgroup definitions — making it impossible to
align results across imputations.

### The stacked-imputation solution

@Rodgers2021 proposed **stacking** the M imputed datasets vertically into
a single data frame of M × n rows, then fitting one model to this combined
data. This circumvents the structural-mismatch problem and yields a single,
coherent, interpretable tree.

The only complication: stacking inflates the nominal sample size by M, so
node-level test statistics are similarly inflated and the tree will
over-split if uncorrected.

### The Stack ÷ M correction

@Sherlock2026 proposed and validated the **Stack ÷ M** correction: use a
significance threshold of `alpha / M` in `ctree_control()`. This is
equivalent to dividing each node's test statistic by M before the pruning
decision, counteracting the inflated sample size. Monte Carlo simulations
under MCAR confirmed that this approach yields:

- **Sub-nominal (conservative) type-I error** (~0.03, 0.02, 0.01 for 10%,
  30%, 50% MCAR, respectively, with M = 30 and n = 9,000)
- **Reduced but acceptable power** compared to permutation-based approaches

The conservatism is a deliberate, acceptable trade-off: it reduces the
chance of spurious splits, which is desirable in exploratory analyses where
interpretability of a single tree is the priority.

---

## Quick start

```{r quickstart, eval = FALSE}
library(ctreeMI)
library(mice)

# 1. Impute your data
imp <- mice(my_data, m = 30, printFlag = FALSE)

# 2. Fit ctree with Stack/M correction
fit <- ctree_stacked(outcome ~ predictor1 + predictor2 + predictor3,
                     data  = imp,
                     alpha = 0.05)

# 3. Inspect and plot
print(fit)
plot(fit)
```

---

## Worked example

We use the `airquality` dataset, introducing artificial missingness to
demonstrate the workflow.

```{r example, eval = FALSE}
library(ctreeMI)
library(mice)

# ── 1. Prepare data ────────────────────────────────────────────────────────
set.seed(42)
aq <- airquality
aq$Ozone[sample(nrow(aq), 25)] <- NA   # 16% missing
aq$Solar.R[sample(nrow(aq), 15)] <- NA  # 10% missing

# ── 2. Multiply impute (M = 20) ───────────────────────────────────────────
imp <- mice(aq, m = 20, printFlag = FALSE)

# ── 3. Fit ctree with Stack/M correction ──────────────────────────────────
fit <- ctree_stacked(
  Ozone ~ Solar.R + Wind + Temp + Month,
  data    = imp,
  m       = 20,
  alpha   = 0.05,
  verbose = TRUE
)
# [ctreeMI] Stacked 20 imputed datasets (n = 153 each; 3060 total rows).
# [ctreeMI] Nominal alpha = 0.0500  |  Stack/M corrected alpha = 0.002500

# ── 4. Inspect results ─────────────────────────────────────────────────────
print(fit)
summary(fit)
plot(fit)
```

---

## Using a list of data frames

If you already have a list of imputed data frames (e.g., from a custom
imputation pipeline), pass them directly:

```{r list-example, eval = FALSE}
# Suppose imp_list is a list of 30 imputed data frames
fit <- ctree_stacked(y ~ x1 + x2 + x3,
                     data  = imp_list,
                     alpha = 0.05)
```

---

## Helper functions

`ctreeMI` exports two utility functions used internally that you may also
find useful directly:

### `stack_imputations()`

Stacks a list of imputed data frames vertically, adding an imputation index
column (`.imp`):

```{r stack, eval = FALSE}
stacked <- stack_imputations(imp_list)
nrow(stacked)         # M * n
table(stacked$.imp)   # equal counts per imputation
```

### `rescale_alpha()`

Computes the adjusted significance threshold:

```{r rescale, eval = FALSE}
rescale_alpha(alpha = 0.05, m = 30)
#> [1] 0.001666667
```

---

## Interpreting results

The object returned by `ctree_stacked()` is a standard `constparty`/`party`
object (from `partykit`) with an additional `"ctreeMI"` class. All
`partykit` methods work:

```{r methods, eval = FALSE}
plot(fit)                          # tree diagram
predict(fit, newdata = test_data)  # predictions
partykit::nodeids(fit, terminal = TRUE)  # terminal node IDs
```

Terminal node estimates are proportions or means computed on the **stacked**
data. Because each observation appears M times (once per imputation),
node-level estimates reflect the average across imputations rather than
inflated counts from a single dataset.

> **Important**: Terminal node sample sizes reported in `print(fit)` and
> `plot(fit)` reflect the **stacked** dataset (M × n). Divide by M to
> obtain the effective per-node sample size in your original data.

---

## Choosing M

There is no universally correct M. General guidance [@Rubin1987;
@VanBuuren2018]:

- **M = 5–10** is sufficient for many applications with low to moderate
  missingness.
- **M = 20–50** is preferable when missingness is substantial (>20%) or
  when precise estimates are needed.
- For the Stack ÷ M correction, larger M increases conservatism (smaller
  corrected alpha). With very large M, the tree may not split at all. In
  practice, M = 20–30 is a reasonable default for most ctreeMI
  applications, consistent with Sherlock et al. (2026).

---

## Citation

If you use `ctreeMI`, please cite the methodological paper and the
underlying `partykit` implementation:

```
Sherlock, P., Mansolf, M., Hofheimer, J., Hockett, C. W., O'Connor, T. G.,
  Roubinov, D., Graff, J. C., Lai, J.-S., Bush, N. R., Wright, R. J., &
  Chiu, Y.-H. M. (2026). Beyond linear risk: A machine learning approach to
  understanding perinatal depression in context.
  Multivariate Behavioral Research, 1-16.
  https://doi.org/10.1080/00273171.2026.2661244

Hothorn, T., Hornik, K., & Zeileis, A. (2006). Unbiased recursive
  partitioning: A conditional inference framework. Journal of Computational
  and Graphical Statistics, 15(3), 651-674.

Hothorn, T., & Zeileis, A. (2015). partykit: A modular toolkit for
  recursive partitioning in R. Journal of Machine Learning Research,
  16, 3905-3909.
```

You can also retrieve the citation from within R:

```{r cite, eval = FALSE}
citation("ctreeMI")
```

---

## References
